---
title: "Practical 2 - PCA"
author: "Kevin Patyk"
date: "11/25/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

-----

# Introduction

In this practical, we will perform singular value decomposition and perform principal component analysis. You will not have to load any packages in advance, as we will solely use base R. The data for today's practical is: `Example1.dat`. 

# SVD and Eigendecomposition

In this exercise, the data of Example 1 from the lecture slides are used. These data have been stored in the file `Example1.dat`.

**1: Use the function `read.table()` to import the data into `R`. Use the function `as.matrix()` to convert the data frame to a matrix. The two features are not centered. To center the two features the function `scale()` with the argument `scale = FALSE` can be used. Give the centered data matrix a name, for instance, `C`.**
```{r}
ex <- read.table("Example1.dat") 

C <- ex %>%
  as.matrix() %>%
  scale(scale = F, center = T)
```

**2: Calculate the sample size N and the covariance matrix S by executing the following R code, where the function `t()` is used to calculate the transpose of `C` and `%*%` is used for matrix multiplication.**
```{r}
#save the number of rows
N <- nrow(C)
#calculate the covariance matrix 
S <- t(C) %*% C/N
```

**3: Use the function `svd()` to apply a singular value decomposition to the centered data matrix.**
```{r}
(svdec <- svd(C))
```

**4: Inspect the three pieces of output, that is, U, D, and V. Are the three matrices the same as on the slides?**

Yes, these are the same as on the slides. Additionally, `$d` is just the single scores rather than a matrix. 

**5: Use a single matrix product to calculate the principal component scores.**
```{r}
U <- svdec$u
D <- matrix(data = c(svdec$d[1], 0, 0, svdec$d[2]), nrow = 2, ncol = 2, byrow = F)

#this can also be done like this: diag(svdec$d)

(components <- U%*%D)
```

**6: Plot the scores on the second principal component (y-axis) against the scores on the first principal component (x-axis) and let the range of the x-axis run from -18 to 18 and the range of the y-axis from -16 to 16.**
```{r}
plot(components[,1], components[,2], ylim = c(-16, 16), xlim = c(-18, 18))
```

**7: Use the function `eigen()` to apply an eigendecomposition to the sample covariance matrix.**
```{r}
(es <- (eigen(S)))
```

**8. Check whether the eigenvalues are equal to the variances of the two principal components. Be aware that the R-base function `var()` takes N - 1 in the denominator, to get an unbiased estimate of the variance.**
```{r}
#scale the component scores
compC <- scale(components, scale = FALSE)
#use matrix multiplication to see if they are equal
round(t(compC) %*% compC/N, 5)
```

They are identical.

**9: Finally, calculate the percentage of total variance explained by each principal component.**
```{r}
round(es$values/sum(es$values)*100, 2)
```

-----

# Principal component analysis

In this exercise, a PCA is used to determine the financial strength of insurance companies. Eight relevant features have been selected: (1) gross written premium, (2) net mathematical reserves, (3) gross claims paid, (4) net premium reserves, (5) net claim reserves, (6) net income, (7) share capital, and (8) gross written premium ceded in reinsurance.

To perform a principal component analysis, an eigen decomposition can be applied to the sample correlation matrix `R` instead of the sample covariance matrix `S`. Note that the sample correlation matrix is the sample covariance matrix of the standardized features. These two ways of doing a PCA will yield different results. If the features have the same scales (the same units), then the covariance matrix should be used. If the features have different scales, then it’s better in general to use the correlation matrix because otherwise the features with high absolute variances will dominate the results.

Loading the sample correlation matrix.
```{r message=FALSE, warning=FALSE}
(cor_mat <- read_csv("Example2.csv")  %>%
  select(-1) %>%
  as.matrix() %>%
  unname())
```

**9: Use R to apply a PCA to the sample correlation matrix.**

An alternative criterion for extracting a smaller number of principal components m than the number of original variables p in applying a PCA to the sample correlation matrix, is the eigenvalue-greater-than-one rule. This rule says that m (the number of extracted principal components) should be equal to the number of eigenvalues greater than one. Since each of the standardized variables has a variance of one, the total variance is p. If a principal component has an eigenvalue greater than one, than its variance is greater than the variance of each of the original standardized variables. Then, this principal component explains more of the total variance than each of the original standardized variables. With a correlation matrix, we can only figure out the eigenvalues and percentages of explained variance, but not the principal components scores because we do not have `U` and `D` matrices. 
```{r}
(res <- eigen(cor_mat))
```

**10: How many principal components should be extracted according to the eigenvalue-greater-than-one rule?**
```{r}
res$values[which(res$values > 1)]
```

According to the eigenvalue-greater-than-one-rule, we should extract 3 principle components.

**11: How much of the total variance does this number of extracted principal components explain?**
```{r}
sum(res$values[1:3])/ sum(res$values)*100
```

This number of principle components explains 89% of the variance.

**12: Make a scree-plot. How many principal components should be extracted according to the scree-plot?**
```{r}
plot(res$values, type = "l")
```

**13: How much of the total variance does this number of extracted principal components explain?**

According to the scree plot, there should be one component because the "elbow" occurs at 2. However, the first principal component by itself only accounts for 58% of the variance, which is not very much. It can also be argued that the "elbow" is at 6 as well. 
```{r}
res$values[1]/sum(res$values)
```

When the second principal component is added to the explained variance, the result is that 76% of the variance is explained. 
```{r}
sum(res$values[1:2])/sum(res$values)
```

With the addition of the third principal component, the variance becomes 89%. This could be a good stopping point, since the number of features would be reduce to 3 (as opposed to the original 8) and still explain nearly 90% of the variance. 
```{r}
sum(res$values[1:3])/sum(res$values)
```

-----

# Lab exercise

In this assignment, you will perform a PCA to a simple and easy to understand dataset. You will use the `mtcars` dataset, which is built into `R`. This dataset consists of data on 32 models of car, taken from an American motoring magazine (1974 Motor Trend magazine). For each car, you have 11 features, expressed in varying units (US units). They are as follows:

* `mpg`: fuel consumption (miles per (US) gallon); more powerful and heavier cars tend to consume more fuel.
* `cyl`: number of cylinders; more powerful cars often have more cylinders.
* `disp`: displacement (cu.in.); the combined volume of the engine’s cylinders.
* `hp`: gross horsepower; this is a measure of the power generated by the car.
* `drat`: rear axle ratio; this describes how a turn of the drive shaft corresponds to a turn of the wheels. Higher values will decrease fuel efficiency.
* `wt`: weight (1000 lbs).
* `qsec`: 1/4 mile time, the cars speed and acceleration.
* `vs`: engine block; this denotes whether the vehicle’s engine is shaped like a ‘V’, or is a more common straight shape.
* `am`: transmission; this denotes whether the car’s transmission is automatic (0) or manual (1).
* `gear`: number of forward gears; sports cars tend to have more gears.
* `carb`: number of carburetors; associated with more powerful engines.

Note that the units used vary and occupy different scales.

First, the principal components will be computed. Because PCA works best with numerical data, you’ll exclude the two categorical variables (`vs` and `am`; columns 8 and 9). You are left with a matrix of 9 columns and 32 rows, which you pass to the `prcomp()` function, assigning your output to `mtcars.pca`. You will also set two arguments, `center` and `scale`, to be `TRUE`. This is done to apply a principal component analysis to the standardized features.:
```{r}
mtcars.pca <- prcomp(mtcars[, c(1:7, 10, 11)],
                     center = TRUE,
                     scale = TRUE)
```

**14: Have a peek at the PCA object with `summary()`.**
```{r}
summary(mtcars.pca)
```

You obtain 9 principal components, which you call PC1-9. Each of these explains a percentage of the total variance in the dataset.

**15: What is the percentage of total variance explained by PC1?**
```{r}
sum_mt <- summary(mtcars.pca)
sum_mt$importance[, 1]
```

PC1 explains 63% of the total variance. 

**16: What is the percentage of total variance explained by PC1, PC2, and PC3 together?**
```{r}
sum_mt$importance[, 3]
```

The total variance explained by PC1, PC2, and PC3 is 92%. 

The PCA object `mtcars.pca` contains the following information:

* the center point or the vector of feature means (`$center`)
* the vector of feature standard deviations (`$scale`)
* the vector of standard deviations of the principal components (`$sdev`)
* the eigenvectors (`$rotation`)
* the principal component scores (`$x`)

**16: Determine the eigenvalues. How many principal components should be extracted according to the eigenvalue-greater-than-one rule?**
```{r}
(evs <- sum_mt$sdev^2)
```

According to the eigenvalue-greater-than-one rule, 2 principal components should be selected.

**17: What is the value of the total variance? Why?**
```{r}
sum(evs)
```

The value of the total variance is 9. For a covariance or correlation matrix, the sum of its eigenvalues equals the trace of the matrix, that is, the sum of the variances of the n variables for a covariance matrix, and n for a correlation matrix.

**18: How much of the total variance is explained by the number of extracted principal components according to the eigenvalue-greater-than-one rule?**
```{r}
sum_mt$importance[, 2]
```

The total of variance explained by the first 2 principal components is 86%. 

Next, a couple of plots will be produced to visualize the PCA solution. You will make a biplot, which includes both the position of each observation (car model) in terms of PC1 and PC2 and also will show you how the initial features map onto this. A biplot is a type of plot that will allow you to visualize how the observations relate to one another in the PCA (which observations are similar and which are different) and will simultaneously reveal how each feature contributes to each principal component.

**19: Use the function `biplot()` with the argument `choices = c(1, 2)` to ask for a biplot for the first two principal components.**
```{r}
biplot(mtcars.pca, choices = c(1,2))
```

You probably notice that a PCA biplot simply merge an usual PCA plot with a plot of loadings. The arrangement is like this:

* Bottom axis: PC1 score.
* Left axis: PC2 score.
* Top axis: loadings on PC1.
* Right axis: loadings on PC2.

In other words, the left and bottom axes are of the PCA plot — use them to read PCA scores of the samples (dots/names). The top and right axes belong to the loading plot — use them to read how strongly each characteristic (vector) influence the principal components.

The axes of the biplot are seen as arrows originating from the center point. Here, you see that the variables `hp`, `cyl`, and `disp` all contribute to PC1, with higher values in those variables moving the observations to the right on this plot. This lets you see how the car models relate to the axes. You can also see which cars are similar to one another. For example, the Maserati Bora, Ferrari Dino and Ford Pantera L all cluster together at the top. This makes sense, as all of these are sports cars.

**20: Make a biplot for the first and third principal components. Especially which brand of car has negative values on the first principal component and positive values on the third principal component?**
```{r}
biplot(mtcars.pca, choices = c(1,3))
```

On the third principal component, the `Merc` vehicles have positive values. On the first principal component, cars like `Toyota` and `Mazda` have negative scores.

**21: Use the function `screeplot()` with the argument `type = 'lines'` to produce a scree-plot. How many principal components should be extracted according to this plot? Why? Is this number in agreement with the number of principal components extracted according to the eigenvalue-greater-than-one rule?**
```{r}
screeplot(mtcars.pca, type = "l")
```

According to the scree plot, the "elbow" is at 3, meaning that 2 principal components should be extracted. This is because the first 2 principal components explain 86% of the variance in the data. This is in agreement with the eigenvalue-greater-than-one rule.

-----

# End of document

-----

```{r}
sessionInfo()
```

